\documentclass{beamer}
\usepackage{hyperref}
\begin{document}

\title{libcxx}
\author{Sebastian Pop and Aditya Kumar}
\institute{SARC: Samsung Austin R\&D Center}
\date{\today}

\frame{\titlepage}

\frame{\frametitle{libcxx-performance analysis}

Abstract: We will discuss the improvements and future work on libcxx. This
includes the improvements on standard library algorithms like string::find and
basic_streambuf::xsgetn. These algorithms were suboptimal and we got huge
improvements after optimizing them. Similarly, we enabled the inlining of
constructor and destructor of std::string. We will present a systematic analysis
of function attributes in libc++ and the places where we added missing
attributes. Finally, we will present a comparative analysis of clang-libc++
vs. gcc-libstdc++ on representative benchmarks


Keywords: libc++, libstdc++, C++, performance analysis, function attributes


  \begin{itemize}
  \item Improving standard library algorithms like string::find, basic_streambuf::xsgetn
  \item inlining the constructor and destructor of std::string
  \item systematic analysis of attributes in the standard library
  \item comparative analysis of clang-libc++ vs. gcc-libstdc++ on representative benchmarks.
  \item clang+libc++ vs. gcc+libstdc++ performance update. Constructor of vector in libstdc++ contains memcpy
    in the final assembly.
  \end{itemize}

  \begin{itemize}
  \item PR21192 - Reading from stdin is 1-2 orders of magnitude slower than using libstdc++
    - implementation of sync\_with\_stdio doesn't disable synchronization \url{http://stackoverflow.com/q/33837349}

  \item PR19708 - std::find is significantly slower than libstdc++
    - using Duff's device
    - unrolling the loop
    - enabling vectorization \url{http://stackoverflow.com/q/2741859}

  \item PR20837 - libc++'s std::sort is O(N^2) in the worst case (instead of O(N*ln(N))).
    - switching to heapsort for large size

  \item PR26886 - libc++'s std::stable\_sort also has a worst-case complexity issue

  \item PR15456 - A faster implementation of std::function is possible

  \item PR16747 and PR21275 - Our unordered\_multimap insert is much slower than libstdc++'s




Benchmark	                                Time (ns)	CPU (ns)	Iterations	Time (ns)	CPU (ns)	Iterations	%(Time)
BM_ConstructIterIter/vector_char/1024	2125	2127	283762	67	67	10205902	3071.641791
BM_ConstructIterIter/vector_size_t/1024	382	383	1830386	223	223	3138652	71.30044843
BM_ConstructIterIter/vector_string/1024	557109	557641	1235	354504	354834	1980	57.15168235

  \item Identifying missing attributes in functions 

  \end{itemize}
}

\end{document}
